Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-07-27 manuscript/manuscript.tex
tex cleanup [ci skip]
added amm citation [ci skip]

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{figures/}}

@@

\newcommand{\versionnumber}{0.3} % TODO: can we extract this number from a git tag?
\newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}

@@

\corr{christoph.wehmeyer@fu-berlin.de}{CW}
%\corr{m.scherer@fu-berlin.de}{MKS}
%\corr{tim.hempel@fu-berlin.de}{TH}
%\corr{simon.olsson@fu-berlin.de}{SO}
%\corr{bhusic@stanford.edu}{BEH}
\corr{frank.noe@fu-berlin.de}{FN}

@@

In addition to publications on theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks},
we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.

@@

Subsequently, we perform TICA~\cite{tica,kinetic-maps} in order to reduce the dimension from the feature space, which typically contains many degrees of freedom, to a lower dimensional space that can be discretized with higher resolution and better statistical efficiency. TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics in the dataset. Here, performing TICA on the backbone torsions at lag time $0.5$ ns yields a four dimensional subspace using a $95\%$ kinetic variance cutoff (note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
The sample density projected onto the first two independent components (ICs) exhibits several maxima (Fig.~\ref{fig:io-to-ck}c).
Discrete jumps between the maxima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-ck}d).
 % time series visualization of these components from the first trajectory shows discrete jumps between these maxima (Fig.~\ref{fig:io-to-ck}d).
We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.

@@

To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test. The CK test compares the right and the left side of the Chapman-Kolmogorov equation
\begin{equation}
\label{eq:ck}
T(k \tau) = T^k(\tau)
\end{equation}
where $T$ is the MSM transition matrix. The left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
Visualizing the full transition probability matrix $T$ is difficult; we therefore coarse-grain $T$ into a smaller number of metastable states before performing the test.
An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot.
% In our case we have multiple gaps and therefore various options to choose from.
For this analysis, we chose 5 metastable states.
%, as it is the largest number that only includes processes which are well above the MSM lag time.
% Given the number of metastable states, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$ using a coarse-grained representation.
The CK test confirms that this is an appropriate choice and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (Fig.~\ref{fig:io-to-ck}f).

@@

\begin{figure}
\includegraphics{figure_3}
\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone:
(a) The reweighted free energy surface projected onto the first two independent components exhibits five minima which
(b) PCCA++ identifies as five metastable states.
(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states, in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
(d) the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$. 
% the most probable state $\mathcal{S}_5$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_4$. 
(e) The Trp-1 SASA autocorrelation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$ (bottom).}
\label{fig:msm-analysis}
\end{figure}

@@

\subsection{Connecting the MSM with experimental data}
MSMs can also be analyzed in the context of experimental observables. Connecting MSM analysis to experimental data can both serve as an accuracy test of our MSM as well as provide a mechanistic interpretation of observed experimental signals.
Since we have both the stationary and dynamic properties of the molecular system encoded in the MSM transition probability matrix, we can compute observables that involve both stationary ensemble averages as well as correlation functions.

As an example, here we look at the fluorescence correlation of Trp-1, since this terminal tryptophan is a realistic experimental observable for our pentapeptide system.
In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each of the original $75$ MSM microstates.
%These values are computed using a 'forward model'.
To approximate the fluorescence signal in our pentapeptide system, we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.
Now that we have an approximation of the fluorescence in each of our MSM states, we can use PyEMMA to compute the fluorescence autocorrelation function (ACF) from our MSM (\ref{fig:msm-analysis}e, upper).
Note how the computed ACF has a very small response (i.e., signal amplitude).

>>>>

\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{figures/}}

@@

\newcommand{\versionnumber}{0.3}
\newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}

@@

\corr{christoph.wehmeyer@fu-berlin.de}{CW}
\corr{frank.noe@fu-berlin.de}{FN}

@@

In addition to publications on theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.

@@

Subsequently, we perform TICA~\cite{tica,kinetic-maps} in order to reduce the dimension from the feature space, which typically contains many degrees of freedom, to a lower dimensional space that can be discretized with higher resolution and better statistical efficiency. TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics in the dataset. Here, performing TICA on the backbone torsions at lag time $0.5$ ns yields a four dimensional subspace using a $95\%$ kinetic variance cutoff (note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
The sample density projected onto the first two independent components (ICs) exhibits several maxima (Fig.~\ref{fig:io-to-ck}c).
Discrete jumps between the maxima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-ck}d).
We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.

@@

To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test. The CK test compares the right and the left side of the Chapman-Kolmogorov equation
\begin{equation}
\label{eq:ck}
T(k \tau) = T^k(\tau)
\end{equation}
where $T$ is the MSM transition matrix. The left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
Visualizing the full transition probability matrix $T$ is difficult; we therefore coarse-grain $T$ into a smaller number of metastable states before performing the test.
An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot.
For this analysis, we chose 5 metastable states.
The CK test confirms that this is an appropriate choice and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (Fig.~\ref{fig:io-to-ck}f).

@@

\begin{figure}
\includegraphics{figure_3}
\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone:
(a) The reweighted free energy surface projected onto the first two independent components exhibits five minima which
(b) PCCA++ identifies as five metastable states.
(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states, in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
(d) the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.
(e) The Trp-1 SASA autocorrelation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$ (bottom).}
\label{fig:msm-analysis}
\end{figure}

@@

\subsection{Connecting the MSM with experimental data}

MSMs can also be analyzed in the context of experimental observables. Connecting MSM analysis to experimental data can both serve as an accuracy test of our MSM as well as provide a mechanistic interpretation of observed experimental signals.
Since we have both the stationary and dynamic properties of the molecular system encoded in the MSM transition probability matrix, we can compute observables that involve both stationary ensemble averages as well as correlation functions.

As an example, here we look at the fluorescence correlation of Trp-1, since this terminal tryptophan is a realistic experimental observable for our pentapeptide system.
In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each of the original $75$ MSM microstates.
To approximate the fluorescence signal in our pentapeptide system, we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.
Now that we have an approximation of the fluorescence in each of our MSM states, we can use PyEMMA to compute the fluorescence autocorrelation function (ACF) from our MSM (\ref{fig:msm-analysis}e, upper).
Note how the computed ACF has a very small response (i.e., signal amplitude).
